### 2016-03-01 doc/paper/paper.Rnw

\documentclass[12pt]{report}
\usepackage[osf]{garamondx}
\usepackage[garamondx,cmbraces]{newtxmath}
\usepackage{marginnote}
\usepackage[top=2cm, bottom=2cm, outer=3cm, inner=3cm, heightrounded, 
            marginparwidth=2cm, marginparsep=0.5cm]{geometry}
\usepackage{setspace}
\usepackage[backend=biber, style=authoryear]{biblatex}
\usepackage{nicefrac}
\usepackage{amsmath}
\usepackage{algorithm}
\usepackage{algorithmicx}
\usepackage{algpseudocode}
\usepackage[acronym,nonumberlist]{glossaries}

\newcommand{\comment}[1]{\marginnote{\textit{#1}}}

@@

\newacronym{BA}{BA}{Barab\'asi-Albert}
\newacronym{SI}{SI}{susceptible-infected}
\newacronym{kSVM}{kSVM}{kernel support vector machine}
\newacronym{nLTT}{nLTT}{normalized lineages-through-time}
\newacronym{SMC}{SMC}{sequential Monte-Carlo}
\newacronym{ABC}{ABC}{approximate Bayesian computation}
\newacronym{MSM}{MSM}{men who have sex with men}

@@

@@

From a theoretical perspective, network structure can profoundly influence the
incidence and prevalence trajectories of an epidemic, in turn affecting the
estimates of quantities such as effective population size. More practically,
contact networks have been explored as tools for curtailing epidemic spread, by
way of interventions targeted to well-connected
nodes~\autocite{little2014using}.

Survey-based studies of sexual networks~\autocite{liljeros2001web,
schneeberger2004scale} have found that sexual contact networks are best
described by a preferential attachment model \autocite[but not all,
see][]{jones2003assessment}. This result has been coroborated by more recent
investigations using phylogenetic methods~\autocite{brown2011transmission}.
Under these models, nodes with a high number of contacts attract new
connections at an elevated rate. Networks produced by preferential attachment
have a power-law degree distribution, meaning that the number of nodes of
degree $k$ is proportional to $k^\gamma$ for some constant $\gamma$. When $2 <
\gamma \leq 3$, the network is referred to as ``scale-free''. The first contact
network model incorporating preferential attachment was introduced by
\textcite{barabasi1999emergence}, and is now referred to as the \gls{BA} model.
Under this model, networks are formed by iteratively adding nodes with $m$ new
edges each. These new edges are joined to existing nodes of degree $k$ with
probability proportional to $k^\alpha$, so that nodes of high degree tend to
attract more connections (in the original paper, only $\alpha = 1$ was
investigated).

@@

Here, we apply kernel-\gls{ABC} to the problem of statistical inference of
contact network parameters from an estimated transmission tree. We then
estimate the parameters of the \gls{BA} model on a variety of simulated and
real data sets.

@@

We simulated three transmission trees, each with 500 tips, under every element
of the Cartesian product of these parameter values: $N$ = 5000, $I$ =
\sett{1000, 2000}, $m$ = \sett{2, 3, 4}, and $\alpha$ = \sett{0, 0.5, 1, 1.5,
2}. The adaptive \gls{ABC} algorithm was applied to each tree with these
priors: $N$ = Uniform(500, 15000), $I$ = Uniform(500, 5000), $m$ = Uniform(1,
5), and $\alpha$ = Uniform(0, 2). Following \textcite{del2012adaptive} and
\textcite{beaumont2009adaptive}, all proposals were Gaussian, with variance
equal to twice the empirical variance of the particles. Because there must be
at least as many nodes in the network as infected nodes, the proposal ratio was
set to zero whenever $I > N$. The algorithm was run with 1000 particles, 5
simulated datasets per particle, and the ``quality'' parameter set to 0.95. We
use the same stopping criterion as \textcite{del2012adaptive}, namely when the
MCMC acceptance rate dropped below 1.5\%. Point estimates for the parameters
were obtained by taking the highest point of an estimated kernel density on the
final set of particles.

@@

We applied our kernel-ABC method to several HIV datasets. Because the \gls{BA}
model generates networks with a single connected component, we specifically
searched for datasets which originated from existing clusters, either
phylogenetically or geographically defined. \textcite{wang2015targeting}
reported on a cohort of acutely infected \gls{MSM} in Beijing, China. Through
phylogenetic analysis, they found that they had ``deeply sampled'' the
local \gls{MSM} network, suggesting that many of the samples could be
epidemiologically related.

We downloaded all sequences associated with each paper from GenBank. Sequences
were multiply aligned using \software{MUSCLE} version 3.8.31, and alignments
were manually inspected with \software{Seaview} version 4.4.2. Phylogenies were
constructed from the nucleotide alignments by approximate maximum likelihood
using \software{FastTree2} version 2.1.7 with the GTR model. Transmission trees
were estimated by rooting and time-scaling the phylogenies by root-to-tip
regression, using a modified version of Path-o-Gen as described
previously~\autocite{poon2015phylodynamic}.

@@

Accuracy of the \gls{kSVM} classifiers varied based on the parameter being
tested. For appropriate choices of the kernel meta-parameters, the classifier
had an $R^2$ above 0.8 for all combinations of tree size and epidemic size. The
$I$ parameter was also classified accurately, with an average $R^2$ of 0.93 for
500-tip trees or 0.70 for 100-tip trees. The $m$ parameter was much harder to
classify, with $R^2$ varying between 0.01 and 0.36 depending on $I$ and the
tree size. Finally, the accuracy of the classifier for $N$ varied widely, from
0.08 with 100 tips and $I = 500$, to 0.82 with 1000 tips and $I = 2000$.
Figure~\ref{fig:crossv} shows the results for $\alpha$; similar figures for the
other parameters can be found in the supplemental materials. Based on
inspection of the cross-validation results, we chose to use the meta-parameters
$\lambda = 0.3$ and $\sigma = 4$ in further analyses, and not to use Sackin's
index or the \gls{nLTT}.

@@

>>>>

\documentclass[12pt]{article}
\usepackage[osf]{garamondx}
\usepackage[garamondx,cmbraces]{newtxmath}
\usepackage{marginnote}
\usepackage[top=2cm, bottom=2cm, outer=3cm, inner=3cm, heightrounded, 
            marginparwidth=2cm, marginparsep=0.5cm]{geometry}
\usepackage{setspace}
\usepackage[backend=biber, style=authoryear, uniquelist=minyear, uniquename=false]{biblatex}
\usepackage[acronym,nonumberlist]{glossaries}
\usepackage{multirow}

@@

\newacronym{BA}{BA}{Barab\'asi-Albert}
\newacronym{SI}{SI}{susceptible-infected}
\newacronym{kSVM}{kSVM}{kernel support vector machine}
\newacronym{nLTT}{nLTT}{normalized lineages-through-time}
\newacronym{SMC}{SMC}{sequential Monte-Carlo}
\newacronym{ABC}{ABC}{approximate Bayesian computation}
\newacronym{MSM}{MSM}{men who have sex with men}
\newacronym{HPD}{HPD}{highest posterior density}

@@

<<setup, include=FALSE>>=
    library(netabc)
    library(xtable)
@

@@

The structure of the contact network underlying a viral epidemic can profoundly
impact the speed and pattern of the epidemic's spread. Network structure can
influence the prevalence trajectory~\autocite{o2010contact} and epidemic
threshold~\autocite{barthelemy2005dynamical}, in turn affecting the estimates
of quantities such as effective population
size~\autocite{goodreau2006assessing}. From a public health perspective,
contact networks have been explored as tools for curtailing epidemic spread, by
way of interventions targeted to well-connected
nodes~\autocite{little2014using}. True contact networks are a challenging type
of data to collect, requiring extensive epidemiological
investigation~\autocite{welch2011statistical}.

Viral sequence data, on the other hand, has become relatively inexpensive and
straightforward to collect on a population level. Due to the high mutation rate
of RNA viruses, epidemiological processes impact the course of viral evolution,
thereby shaping the intra-host viral
phylogeny~\autocite{drummond2003measurably}. The term ``phylodynamics'' was
coined to describe this interaction, as well as the growing family of inference
methods to estimate epidemiological parameters from viral
phylogenies~\autocite{grenfell2004unifying}. These methods have revealed
diverse properties of local viral outbreaks, from basic reproductive
number~\autocite{stadler2011estimating}, to the degree of
clustering~\autocite{hughes2009molecular}, to the elevated transmission risk
during acute infection~\autocite{volz2012simple}. On the other hand, inference
of structural network parameters has to date been limited. However, it has been
shown that network structure has a tangible impact on phylogeny
shape~\autocite{leventhal2012inferring, colijn2014phylogenetic,
goodreau2006assessing, robinson2013dynamics}, suggesting that statistical
inference might be possible~\autocite{welch2011statistical}.

Survey-based studies of sexual networks~\autocite{liljeros2001web,
schneeberger2004scale} have found that sexual contact networks are best
described by a preferential attachment model \autocite[although there has been
some disagreement, see][]{jones2003assessment}. Under these models, nodes with
a high number of contacts attract new connections at an elevated rate. Networks
produced by preferential attachment have a power-law degree distribution,
meaning that the number of nodes of degree $k$ is proportional to $k^\gamma$
for some constant $\gamma$. When $2 < \gamma \leq 3$, the network is referred
to as ``scale-free''. The first contact network model incorporating
preferential attachment was introduced by \textcite{barabasi1999emergence}, and
is now referred to as the \gls{BA} model. Under this model, networks are formed
by iteratively adding nodes with $m$ new edges each. These new edges are joined
to existing nodes of degree $k$ with probability proportional to $k^\alpha$, so
that nodes of high degree tend to attract more connections (in the original
paper, only $\alpha = 1$ was investigated).

Previous work offers precedent for the possibility of statistical inference of
structural network parameters. \textcite{britton2002bayesian} develop a
Bayesian approach to estimate the edge density in an Erd\H{o}s-R\'enyi
network~\autocite{erdos1960evolution} given observed infection dates, and
optionally recovery dates. Their approach was later extended by
\textcite{groendyke2011bayesian} and applied to a much larger data set of 188
individuals. \textcite{brown2011transmission} fit a preferential attachment
model to a phylogenetically estimated transmission network relating 60\% of HIV
infected \gls{MSM} in the United Kingdom. The transmission network is a
subgraph of the contact network which includes only those edges which have
already led to a new infection.

@@

Here, we apply kernel-\gls{ABC} to the problem of statistical inference of
contact network parameters from an estimated transmission tree, using the tree
kernel developed by \textcite{poon2013mapping}. We then estimate the parameters
of the \gls{BA} model on a variety of simulated and real data sets.

@@

We simulated three transmission trees, each with 500 tips, under every element
of the Cartesian product of these parameter values: $N$ = 5000, $I$ =
\sett{1000, 2000}, $m$ = \sett{2, 3, 4}, and $\alpha$ = \sett{0.0, 0.5, 1, 1.5}.
The adaptive \gls{ABC} algorithm was applied to each tree with these priors:
$m \sim$ Uniform(1, 5), $\alpha \sim$ Uniform(0, 2), and $(N, I)$ jointly
uniform on the triangular region \{$500 \leq N \leq 15000$, $500 \leq I \leq
 5000$, $I \leq N$\}. Following \textcite{del2012adaptive} and
\textcite{beaumont2009adaptive}, all proposals were Gaussian, with variance
equal to twice the empirical variance of the particles. The algorithm was run
with 1000 particles, 5 simulated datasets per particle, and the ``quality''
parameter set to 0.95. We use the same stopping criterion as
\textcite{del2012adaptive}, namely when the MCMC acceptance rate dropped below
1.5\%. Point estimates for the parameters were obtained by taking the highest
point of an estimated kernel density on the final set of particles.

@@

We applied our kernel-ABC method to several HIV datasets. Because the \gls{BA}
model generates networks with a single connected component, we specifically
searched for datasets which originated from existing clusters, either
phylogenetically or geographically defined. Characteristics of the datasets we
investigated are given in Table~\ref{tab:data}.

\begin{table}
  \centering
  \begin{tabular}{ccccc}
    Reference & Sequences ($n$) & Location & Risk group & Gene \\
    \hline
    \autocite{wang2015targeting} & 173 & Beijing, China & MSM & \textit{pol} \\
    \autocite{cuevas2009hiv} & 287 & Basque Country, Spain & mixed & \textit{pol} \\
    \autocite{novitsky2013phylogenetic} & \multirow{2}{*}{201^*} &
    \multirow{2}{*}{Mochudi, Botswana} & \multirow{2}{*}{mixed} &
    \multirow{2}{*}{\textit{env}} \\ \autocite{novitsky2014impact} \\
    \hline
  \end{tabular}
  \caption{Characteristics of published datasets investigated with kernel-ABC.
  Acronyms: MSM, men who have sex with men. $^*$ Sub-sampled from 1299.}
  \label{tab:data}
\end{table}

We downloaded all sequences associated with each study from GenBank. Sequences
were multiply aligned using \software{MUSCLE} version 3.8.31
\autocite{edgar2004muscle}, and alignments were manually inspected with
\software{Seaview} version 4.4.2 \autocite{gouy2010seaview}. Phylogenies were
constructed from the nucleotide alignments by approximate maximum likelihood
using \software{FastTree2} version 2.1.7 with the GTR model. Transmission trees
were estimated by rooting and time-scaling the phylogenies by root-to-tip
regression, using a modified version of Path-o-Gen as described
previously~\autocite{poon2015phylodynamic}.

@@

<<classifiers, include=FALSE>>=
    get.kernel.data <- function (param) {
        md <- collect.metadata(paste0('../../simulations/kernel-', param, '/classifier/*'))
        md <- subset(md, rbf_variance == 4 & decay_factor == 0.3)
        if (param == 'alpha') {
            md <- subset(md, m == 2)
        }
        d <- setDT(collect.data(rownames(md)))
        if (param %in% c('alpha', 'm')) {
            d[,list(rsquared=mean(rsquared)), by=c('nsimnode', 'ntip')]
        } else if (param == 'I') {
            d[,list(rsquared=mean(rsquared)), by=ntip]
        } else {
            d[,list(rsquared=mean(rsquared)), by=c('I', 'ntip')]
        }
    }
    kernel.alpha <- get.kernel.data('alpha')
    kernel.m <- get.kernel.data('m')
    kernel.I <- get.kernel.data('I')
    kernel.N <- get.kernel.data('N')
@

Accuracy of the \gls{kSVM} classifiers varied based on the parameter being
tested. For appropriate choices of the kernel meta-parameters, the classifier
had an $R^2$ above 
    \Sexpr{kernel.alpha[,min(rsquared %/% 0.05 * 0.05)]} 
for all combinations of tree size and epidemic size. The $I$ parameter was also
classified accurately, with an average $R^2$ of 
    \Sexpr{kernel.I[ntip == 500, round(mean(rsquared), 2)]} 
for 500-tip trees or 
    \Sexpr{kernel.I[ntip == 100, round(mean(rsquared), 2)]} 
for 100-tip trees. The $m$ parameter was much harder to classify, with $R^2$
varying between 
    \Sexpr{kernel.m[,round(min(rsquared), 3)]}
and
    \Sexpr{kernel.m[,round(max(rsquared), 2)]}
depending on $I$ and the tree size. Finally, the accuracy of the classifier for
$N$ varied widely, from 
    \Sexpr{kernel.N[,round(min(rsquared), 2)]}
to
    \Sexpr{kernel.N[,round(max(rsquared), 2)]}.
Figure~\ref{fig:crossv} shows the results for $\alpha$; similar figures for the
other parameters can be found in the supplemental materials. Based on
inspection of the cross-validation results, we chose to use the meta-parameters
$\lambda = 0.3$ and $\sigma = 4$ in further analyses, and not to use Sackin's
index or the \gls{nLTT}.

@@

Across replicates and other model parameters, the mean absolute difference
between the estimated and true values of $\alpha$ was X, with a standard error
of X.

Figure~\ref{fig:abcpt} shows point estimates for the simulations with $m = 2$.
The results for the other values of $m$ are similar and can be found in the
supplemental materials.

The marginal approximated posterior distributions for one simulated dataset are
shown in Figure~\ref{fig:abcex}.

\begin{figure}
    \includegraphics{abc-point-est}
    \caption{Point estimates of \gls{BA} model parameters obtained by running
    kernel-ABC on simulated phylogenies. Dotted lines indicate the true values,
    and limits of the $y$-axes are the regions of uniform prior density.}
    \label{fig:abcpt}
\end{figure}

\begin{figure}
    \includegraphics{abc-example}
    \caption{Marginal estimated posterior distributions obtained with ABC on a
    single simulated dataset. Dotted lines and shaded polygon indicate true
    values.}
    \label{fig:abcex}
\end{figure}

\subsection*{Real data}

### 2016-03-01 doc/thesis/results.tex

\subsection{Kernel-SVM classifiers}

Our first simulation-based experiment was designed to determine which network
model parameters were potentially estimable by kernel-\gls{ABC}, and to find
the best set of kernel meta-parameters to carry through to future experiments.
We simulated trees under three distinct values of each parameter of the
\gls{BA} model, and calculated the cross-validation accuracy of a
kernel-\gls{SVM} classifier. Many of the figures in this section pertain to the
attachment power parameter $\alpha$ only, but equivalent figures for the other
three parameters can be found in the supplemental materials.

Trees simulated under different values of \gls{alpha} are visibly quite
distinctive (Figure~\ref{fig:patrees}). In particular, higher values of
\gls{alpha} result in networks with a small number of highly connected nodes
(see Figure~\ref{fig:pabounds}) which, once infected, are likely to transmit to
many other nodes. This results in a more unbalanced, ladder-like structure in
the phylogeny.

Kernel-\gls{PCA} projections show that all three \gls{alpha} values are well
separated from each other under several different sampling scenarios
(Figure~\ref{fig:pakpca}). For smaller trees, though, it becomes harder to
visually disinguish \gls{alpha} = 0.5 from \gls{alpha} = 1.0 using the first
two principal components.

With suitably chosen meta-parameters, the accuracy of the kernel-\gls{SVM}
classifier for \gls{alpha} was very high under a variety of prevalence and
sampling scenarios (Figure~\ref{fig:pacrossv}). Accuracy was highest, with
$R^2$ values above 0.95, for the largest trees and complete sampling (bottom
center panel). However, even for trees of size 100 sampled from an epidemic on
2000 nodes, the $R^2$ was above 0.8 (top right panel). In all cases, the
accuracy of a Sackin's index-based classifier was also quite high, at about
0.75. The fact that Sackin's index is informative of \gls{alpha} was
unsurprising, given the clear differences in tree balance observed under
different \gls{alpha} values.

>>>>

\subsection{Separable parameters}

Our first simulation-based experiment was designed to determine which network
model parameters were potentially estimable by kernel-ABC, and to find the best
set of kernel meta-parameters to carry through to future experiments. For each
parameter of the \gls{BA} model, we simulated trees under three distinct
values, and computed the cross-validation accuracy of a kernel-\gls{SVM}
classifier for the parameter.

In Figure~\ref{fig:patrees}, we show that trees simulated under different
values of \gls{alpha} are visibly quite distinctive. In particular, higher
values of \gls{alpha} result in networks with a small number of highly
connected nodes (see Figure~\ref{fig:pabounds}) which, once infected, are
likely to transmit to many other nodes. This results in a more unbalanced,
ladder-like structure in the phylogeny. Kernel-\gls{PCA} projections show that
all three \gls{alpha} values are well separated from each other under several
different sampling scenarios (Figure~\ref{fig:pakpca}). With smaller trees,
it becomes harder to visually disinguish \gls{alpha} = 0.5 from \gls{alpha} =
1.0 using the first two principal components.

With suitably chosen meta-parameters, the accuracy of the kernel-\gls{SVM}
classifier for \gls{alpha} was very high under a variety of prevalence and
sampling scenarios (Figure~\ref{fig:pacrossv}). Accuracy was highest, with
$R^2$ values above 0.95, for the largest trees and complete sampling (bottom
center panel). However, even for trees of size 100 sampled from an epidemic on
2000 nodes, the $R^2$ was above 0.8 (top right panel). In all cases, the
accuracy of a Sackin's index-based classifier was also quite high, at about
0.75. An \gls{SVM} classifier using only the \gls{nltt}
statistic~\autocite{janzen2015approximate} was slightly worse than Sackin's
index, with an $R^2$ of X.

The fact that Sackin's index is informative of \gls{alpha} was
unsurprising, given the clear differences in tree balance observed under
different \gls{alpha} values.
